---
title: "Initial chapter two code"
author: "siefeldin sobih"
date: "`r Sys.Date()`"
output:
  
  html_document: 
    toc: true
    number_sections: true
    keep_md: true
    code_folding: hide
    code_download: true
    theme: cerulean
  pdf_document: 
    latex_engine: xelatex
    toc: true
    number_sections: true
header-includes:
  - |
    ```{=latex}
    \usepackage{fvextra}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
      breaksymbolleft={}, 
      showspaces = false,
      showtabs = false,
      breaklines,
      commandchars=\\\{\}
---
\newpage
```{r setup, include = FALSE}
# set up global R options
options(digits = 3)

# set up knitr global chunk options
knitr::opts_chunk$set(
  fig.width = 8,
  fig.path = "figures/",
  echo = TRUE,
  eval = TRUE,
  results = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  tidy = "styler")
```
\newpage
# Set Seed
```{r Setting Seed}
set.seed(123568)
```
\newpage
# Loading libraries
```{r Essential Libraries}
library(qiime2R)
library(rmarkdown)
library(dplyr)
library(phyloseq)
library(vegan)
library(data.table)
library(readr)
library(ggplot2)
library(decontam)
library(microbiome)
library(biomeUtils)
library(tidyverse)
library(lme4)
library(ANCOMBC)
library(data.table)
library(lme4)
library(readr)
library(car)
library(MuMIn)
library(MASS)
library(mgcv)
library(performance)
library(devtools)
library(pairwiseAdonis)
library(MuMIn)
library(emmeans)
library(cowplot)
library(patchwork)
suppressPackageStartupMessages(library(microViz))
library(lintr)
library(forecast)
```
\newpage
# Step 1: Importing, Editing, and Exporting Metadata and Manifest files
```{r 1 Importing, Editing, and Exporting Metadata and Manifest files, message = FALSE, results = FALSE}
#Upload metadata and manifest (used in qiime2) file
#Manifest IDs are used to match Metadata ID to have equal IDs in the analysis
manifest_2016_pre <- read.delim("pre_analysis_prep/2016_manifest_rockiguana_V2",
                                sep = "\t", header = TRUE)
metadata_2016_pre <- read.csv("pre_analysis_prep/metadata_2016.csv", header = TRUE,
                              sep = ",")
sum(metadata_2016_pre$sample.id %in% manifest_2016_pre$sample.id)#255 metadata ID found in Manifest ID
#Check for what 2 missing IDs were found
setdiff(manifest_2016_pre$sample.id, metadata_2016_pre$sample.id)#F11 and Undetermined (Missing sample IDs)
#Only keep matching IDs found in manifest ID column
metadata2016_qiime2_initial <- metadata_2016_pre %>% filter(sample.id %in% manifest_2016_pre$sample.id)
#Only keep matching IDs found in Metadata ID column (not important for qiime2 analysis)
manifest2016_qiime2_final <- manifest_2016_pre %>% filter(sample.id %in% metadata_2016_pre$sample.id)
write.table(manifest2016_qiime2_final, "manifest2016_qiime2_final.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
#Create a new variable for metadata editing and further downstream analysis
metadata_qiime2<- metadata2016_qiime2_initial
#Any empty space is NA
metadata_qiime2[metadata_qiime2 == ""] <- NA
#Create new Reproductive Season Categorical variable from month column
metadata_qiime2$reproductive_season <- case_when(metadata_qiime2$month == "may" ~ "Breeding", metadata_qiime2$month == "june" ~ "Oviposition",metadata_qiime2$month == "september" ~ "Post-Breeding")
#Check for unique observations in each column
unique(metadata_qiime2$species);unique(metadata_qiime2$sex);unique(metadata_qiime2$tourism.level);unique(metadata_qiime2$repro_score);unique(metadata_qiime2$visited)
#Transform each character column / categorical to factor
metadata_qiime2 <- metadata_qiime2 %>% mutate(across(c("sample.id","iguana.id","unique.id","sample.type","species","site","sex","tourism.level","visited","year","date","month","repro_month","repro_score","reproductive_season"), as.factor))
#Transform the rest of the columns to numeric
metadata_qiime2 <- metadata_qiime2 %>% mutate(across(where(is.character), as.numeric))
#Calculate body mass index - referenced in thesis
metadata_qiime2$body_mass_index<- as.numeric(metadata_qiime2$mass) / as.numeric(metadata_qiime2$svl)^2
#Check ranges for all numerical columns
summary(metadata_qiime2[sapply(metadata_qiime2, is.numeric)])
#Quick check on dataframe structure
str(metadata_qiime2)
```
\newpage
# Step 2: Checking for variables with the most missing data
```{r 2 Missing data count, message = TRUE}
#Checking for columns with high NAs
na_in_metadata <- metadata_qiime2[metadata_qiime2$sample.type=='sample',]

na_summary <- data.frame(column = character(), percent_missing = numeric(), stringsAsFactors = FALSE)

for (col_name in colnames(na_in_metadata)) {
  percent_na <- sum(is.na(na_in_metadata[[col_name]])) / nrow(na_in_metadata) * 100
  na_summary <- rbind(na_summary, data.frame(
    column = col_name,
    percent_missing = percent_na))}

head(arrange(na_summary, desc(percent_missing)),8) #The columns with more than 50% NA

metadata_final <- metadata_qiime2

readr::write_tsv(metadata_final, "analysis_input/metadata_2016_final.tsv")
```
\newpage
# Step 3: Creating Phyloseq object
```{r 3 Phyloseq object, message=TRUE}
ps_16<- qza_to_phyloseq(
    features= "analysis_input/feature2016table.qza",
    tree = "analysis_input/rooted2016-tree.qza",
    taxonomy = "analysis_input/classified2016taxonomy.qza",
    metadata = 'analysis_input/metadata_2016_final.tsv')
ps_16
```
\newpage
# Step 4: Pre-processing phyloseq object
```{r 4 Chloroplast and Mitochondria removal, message=TRUE}
#Check if any sample name has ZYM (mock communities)
sum(grepl("ZYM", sample_names(ps_16)))
#Mitochondria filtration 
"Mitochondria" %in% tax_table(ps_16)[, "Family"]
ps_16_nomitochondria <- subset_taxa(ps_16, Family != "Mitochondria")
ntaxa(ps_16_nomitochondria)#Lost 1048 -> 7314 left
"Mitochondria" %in% tax_table(ps_16_nomitochondria)[, "Family"]
#Chloroplast filtration:
"Chloroplast" %in% tax_table(ps_16_nomitochondria)[, "Order"]
ps_16_nomitono_chloro <- subset_taxa(ps_16_nomitochondria, Order != "Chloroplast")
ntaxa(ps_16_nomitono_chloro)#7285 after removal of 29 chloroplast classification

"Chloroplast" %in% tax_table(ps_16_nomitono_chloro)[, "Order"]
"Mitochondria" %in% tax_table(ps_16_nomitono_chloro)[, "Family"]
```
\newpage
# Step 5: Decontamination and control removal of phyloseq
```{r Step 5 Decontamination and Control removal}
sample_data(ps_16_nomitono_chloro)$is.neg <- sample_data(ps_16_nomitono_chloro)$sample.type == "control"
contamdf_prev <- isContaminant(ps_16_nomitono_chloro, method="prevalence", neg="is.neg")
table(contamdf_prev$contaminant)#Total of 27 contaminants

contamdf_prev05 <- isContaminant(ps_16_nomitono_chloro, method="prevalence", neg="is.neg", threshold=0.5)
table(contamdf_prev05$contaminant)#Found 82 contaminants.

#Extract names of non-contaminated data 
not_contaminant <- row.names(contamdf_prev05[contamdf_prev05[,"contaminant"] == "FALSE",])
#Retain only non-contaminated taxa
ps_decontaminated <- prune_taxa(not_contaminant, ps_16_nomitono_chloro)
ntaxa(ps_decontaminated) # 7203 - removed 82

# remove negative controls (control samples)
ps_decontaminated <- prune_samples(sample_data(ps_decontaminated)$sample.type != "control", ps_decontaminated)
table(sample_data(ps_decontaminated)$sample.type) #18 control removed - 237 samples left

ps_decontaminated_filtered_tax <- subset_taxa(ps_decontaminated, !is.na(Phylum))#Phylum not NA
ps_decontaminated_filtered_tax #No change
```
\newpage
# Step 6: Filter features not seen at least 10 times in each sample
```{r Step 6 Low abundance}
# New approach: Filter taxa with fewer than 10 sequences from individual samples
otu_mat <- as(otu_table(ps_decontaminated_filtered_tax), "matrix")
otu_mat[otu_mat < 10] <- 0 #Filter ASVs PER sample that don't show up 10 times
taxa_to_keep <- rownames(otu_mat)[rowSums(otu_mat) > 0] #This removes any ASV that doesn't have atleast 10 reads in one sample
ps_decontaminated_filtered_tax_kept <- prune_taxa(taxa_to_keep, ps_decontaminated_filtered_tax)
ps_decontaminated_filtered_tax_kept #3960 taxa kept --> Lost 3243 taxa
```
\newpage
# Step 7: Filter samples for sequencing depth
```{r 7 Sequencing depth}
#Remove samples with fewer than 1000 reads
ps_decontaminated_filtered_tax_kept_filteredsample <- prune_samples(sample_sums(ps_decontaminated_filtered_tax_kept) >= 1000, ps_decontaminated_filtered_tax_kept)
nsamples(ps_decontaminated_filtered_tax_kept_filteredsample)#No Change

ps_ch2<-ps_decontaminated_filtered_tax_kept_filteredsample#Original PS preserved

ps_ch2_temp_meta <- data.frame(sample_data(ps_ch2))

#Dataframe exportation to apply metadata transformation that was lost during phyloseq conversion
ps_ch2_temp_meta <- ps_ch2_temp_meta %>% mutate(across(c("iguana.id","unique.id","sample.type","species","site","sex","tourism.level","visited","year","date","month","repro_month","repro_score","reproductive_season"), as.factor))

sample_data(ps_ch2) <- sample_data(ps_ch2_temp_meta)#Replace metadata for correct factorial transformation

#Version for non Z-transformed covarites 
ps_ch2_noZscore<- ps_ch2 #Non-continuous transformed PS object for Visualization
#Version for Z-transformation
ps_ch2_zscore<- ps_ch2 #Non-continuous transformed PS object for Visualization

#Z-score transformation for all continuous predictors that will be used downstream based on previous runs
ps_ch2_temp_meta <- ps_ch2_temp_meta %>% mutate(across(c(svl, tail, mass, glucose, true.triglycerides, testosterone, corticosterone, bacterial.killing.ability,glycerol,true.triglycerides,droms,antioxidants.index,body_mass_index,progesterone,e2), ~as.numeric(scale(.))))

#Thenn Replace metadata in ps_ch2
sample_data(ps_ch2_zscore) <- sample_data(ps_ch2_temp_meta) #Z-score transformed PS Object for downstream analysis
```
\newpage
# Step 8: Data Explorations (Ordination, Inflation factors, LMM)
```{r}
#First pass the phyloseq object to investigate any issues with taxonomy
ps_relative<- ps_ch2_zscore
#Validate Taxonomy and Check for errors in Phyloseq object
phyloseq_validate(ps_relative, remove_undetected = FALSE,min_tax_length = 4,verbose = TRUE)#None found
ps_relative <- ps_relative %>% tax_fix()#No fixes applied, phyloseq ready for further downstream analysis
#Faith calculation from phyloseq and rooted tree
ps_faith<- calculatePD(ps_relative, justDF = TRUE, include_root = TRUE)


#How does the factors look like with only progesterone and e2 in the data?
ps_faith_prog_only <- ps_faith[!is.na(ps_faith$e2), ]
ps_faith_prog_only<-droplevels(ps_faith_prog_only)
str(ps_faith_prog_only)
unique(ps_faith_prog_only$sex)
nrow(ps_faith_prog_only)#95 out of 237! (Lost 142 samples)

#Checking for collinear variables
#Steps that I took after VIF produced an error:
# site and sex produce "there are aliased coefficients in the model" removed both and re-ran the model.
# progesterone and e2 are PERFECTLY collinear with reproductive score
# progesterone and e2 are HIGHLY collinear with reproductive season (14GVIF/1.94), glycerol(7.44/2.73), true.triglycerides(7.94/2.82), and droms (5.12/2.26) AFTER the removal of Reproductive score.

#Variance inflation, ran multiple times, noted above results, RMD doesn't let it run if it produces errors
vif(lm(PD ~  species + tourism.level + reproductive_season + glucose  + testosterone + repro_score + corticosterone + bacterial.killing.ability + glycerol + true.triglycerides + droms + antioxidants.index + body_mass_index, data = ps_faith))

ps_faith$BOXPD <- forecast::BoxCox(ps_faith$PD, lambda = "auto")

#Checking faith distribution for PD model below
hist(ps_faith$PD);hist(log(ps_faith$PD));hist(sqrt(ps_faith$PD));hist(ps_faith$BOXPD)

#Testing assumptions of normality
shapiro.test(ps_faith$PD)
qqnorm(ps_faith$PD); qqline(ps_faith$PD)

shapiro.test(sqrt(ps_faith$PD))
qqnorm(sqrt(ps_faith$PD)); qqline(sqrt(ps_faith$PD))

shapiro.test(log(ps_faith$PD))
qqnorm(log(ps_faith$PD)); qqline(log(ps_faith$PD))

shapiro.test(ps_faith$BOXPD)
qqnorm(ps_faith$BOXPD); qqline(ps_faith$BOXPD)

# Checking for LMM for every variable in the model
#Despite BOXCOX passing shapiro significance test, it breaks homoscedasticity assumption making P values null, and it has a high AIC (727)
#Reverted to sqrt for low 
test_PD_proe2<- lmerTest::lmer(log(PD) ~  species + tourism.level  + reproductive_season + glucose + testosterone + repro_score + corticosterone  + bacterial.killing.ability + glycerol + true.triglycerides + droms + antioxidants.index + body_mass_index + progesterone + e2 + (1|unique.id), data = ps_faith)

summary(test_PD_proe2);plot(test_PD_proe2);AIC(test_PD_proe2)#121 - Only species as significant
check_heteroscedasticity(test_PD_proe2)#Model assumption kept with log not raw
Anova(test_PD_proe2)


test_PD_NOproe2<- lmerTest::lmer(log((PD)) ~  species + tourism.level  + reproductive_season + glucose + testosterone + repro_score + corticosterone  + bacterial.killing.ability + glycerol + true.triglycerides + droms + antioxidants.index + body_mass_index + (1|unique.id), data = ps_faith)

summary(test_PD_NOproe2);plot(test_PD_NOproe2);AIC(test_PD_NOproe2)#410
check_heteroscedasticity(test_PD_NOproe2)#model assumption broken EVEN with reproductive score tested and removed
Anova(test_PD_NOproe2)

# Testing Ordination
ps_ordination_test <- ps_relative
ps_ordination_test_df <- data.frame(sample_data(ps_ordination_test))

ps_ordination_test_df %>% dplyr::select(-c(e2, progesterone))

ps_ordination_test_df <- ps_ordination_test_df %>%
  dplyr::select(species, tourism.level, reproductive_season, glucose, testosterone, repro_score, corticosterone,
                bacterial.killing.ability, glycerol,droms, true.triglycerides, droms, antioxidants.index) %>%
  tidyr::drop_na()

ps_ordination_test_df <- ps_ordination_test_df %>% mutate(across(c("species", "tourism.level", "reproductive_season", "repro_score"), as.factor))

ps_ordination_test <- prune_samples(sample_names(ps_ordination_test) %in% rownames(ps_ordination_test_df), ps_ordination_test)

sample_data(ps_ordination_test) <- sample_data(ps_ordination_test_df)

length(sample_names(ps_ordination_test))#188 without any NA - Lost 49 samples

#Ordination transformed BC for exploration
# Log1p transforms using log(1 + x), safe for zero values
ps_ordination_test <- transform_sample_counts(ps_ordination_test, function(x) log1p(x))
ps_ordination_test_ordination<- ordinate(ps_ordination_test, method = "PCoA", distance = "bray", formula = NULL)

#Transformed Bray Curtis visualization
test_ord_1<- plot_ordination(ps_ordination_test, ps_ordination_test_ordination, color = "species") +
  geom_point(size = 2) +
  labs(color = "Sub-species")+
  theme_minimal() +
  stat_ellipse(type = "t", linetype = "solid") +
  theme(
    legend.title = element_text(size = 16),
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 16),legend.text = element_text(size = 16)
  )

test_ord_2<-plot_ordination(ps_ordination_test, ps_ordination_test_ordination, color = "tourism.level") +
  geom_point(size = 2) +
  labs(color = "Tourism Category")+
  theme_minimal() +
  stat_ellipse(type = "t", linetype = "solid") +
  theme(
    legend.title = element_text(size = 16),
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 16),legend.text = element_text(size = 16)
  )

test_ord_3<-plot_ordination(ps_ordination_test, ps_ordination_test_ordination, color = "reproductive_season") +
  geom_point(size = 2) +
  labs(color = "Reproductive Season")+
  theme_minimal() +
  stat_ellipse(type = "t", linetype = "solid") +
  theme(
    legend.title = element_text(size = 16),
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 16),legend.text = element_text(size = 16)
  )

test_ord_4<-plot_ordination(ps_ordination_test, ps_ordination_test_ordination, color = "repro_score") +
  geom_point(size = 2) +
  labs(color = "Reproductive Score")+
  theme_minimal() +
  stat_ellipse(type = "t", linetype = "solid") +
  theme(
    legend.title = element_text(size = 16),
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 16),legend.text = element_text(size = 16)
  )

test_ordination_combined <- test_ord_1 + test_ord_2 +test_ord_3 +test_ord_4 + plot_annotation(tag_levels = 'A')
ggsave("test_ordination_com.png", plot = test_ordination_combined, width = 16, height = 6, dpi = 300)
```

